---
title: "Plane Crash Analysis: Does your seat matter?"
subtitle: "Inferential Statistics 2025/2026"
author: "Barbato Alberto, Naldoni Martina"
date: "`r Sys.Date()`"
output: 
  pdf_document: 
    toc: true
    fig_height: 6
    number_sections: true
editor_options: 
  markdown: 
    wrap: 80
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Introduction

The Aviation industry is one of the safest in the world. The systems that it has in place to learn lessons from accidents and make sure they never happen again is well-established and highly regarded. This has made traveling by plane the safest way to travel. 
https://flyfright.com/plane-crash-statistics/#:~:text=Based%20on%20statistics%20from%202015,unharmed%2C%20injured%2C%20or%20killed

But perhaps due to these high safety standards, when an aircraft accident happens it makes headlines all over the world. Anxious passengers fear it's going to happen to them as well and they can't help but ask themselves if there is something they could do to have a safer flight. 

One of the most frequent question that gets asked is : is there a part of the plane that is "safer" than other parts? Can the seating location make a difference in an accident? 
There are many articles where experts in the field give their opinions, and most answer that "yes, there are areas that give a higher chance of survival in case of an aircraft accident". 
But there are also studies that suggest that there is no safest seats on an airplane. 

So, who is right? We are going to try to answer this question using statistics.

# Vocabulary

Before starting the analysis, the meaning of some terms must be clarified 

 - Flight : A flight a trip made by an aircraft that connects two airports. It is identified by a number. (Example : AirFrance 225 is the name of the regular service from New Deli to Paris ) 
 
 - Accident : An Accident is an occurrence where a person is fatally or seriously injured, the aircraft sustains significant structural damage, or the aircraft goes missing.
 
 - Incident : An incident is a dangerous situation where no one is seriously hurt and the plane isn't badly damaged.
 
 - Crash : A crash is a type of accident where an aircraft strikes the ground, water, or an obstacle with enough force to cause severe damage or total destruction.
 
 - Accident flight : when a serious accident happens on a flight, usually the companies save the flight number to refer to that flight, and change the flight number for new flights. (Example : Air France 447 was the name of the connection from Rio to Paris, after the 2009 accident the flight became Air France 445)
 
 - Serious injury : Any injury requiring more than 48 hours of hospitalization or involving broken bones (other than fingers/toes)
 
 - Survivability : an accident is defined as survivable if the forces are within human limits and if the structure remained substantially intact 
 
 - CREEP factors : factors that influence the survivability of an accident. They are
    - Container : amount of cabin deformation 
    - Restraints : analysis of the seating structure
    - Environment : presence of lethal contact points
    - Energy Absorption : G-load mitigation by fuselarge or landing gear
    - Post Crash : Other factors (We considered : airport proximity, presence of fire, daytime, phase of flight)

# Perspective 

We are studying if the seating in an airplane has an effect on the survivability of an aircraft accident. To study this, we need to look at all aircraft accidents, then rule out the accidents where every passenger survived, and every accident in which every passenger died. This is an extremely narrow data set of accidents. Of these accidents, we gathered the seatings arrangements and survivor seating maps for 47 crashes. 

If 47 crashes look like a lot, we should consider the greater perspective of air travel safety in general.

https://flyfright.com/plane-crash-statistics/#tve-jump-18c020d9166
reported these figures studying the US General aviation data between 2015 and 2020 : 
 
 - $\frac{1}{260256}$ :  chance of boarding any flight and it being an accident flight
 
 - $\frac{1}{6,864,250}$ : chance of being on a plane involved in an accident that results in at least 1 fatality (possible case study of this study)

 - $\frac{1}{816,545,929}$ chance of you specifically, dying in a plane crash
 


 


```{r}

data <- read.csv("AllCREEP_cleaned.csv")
# trasform the variables that are categorical into factors
data$fonte <- as.factor(data$fonte)
data$PhaseOfFlight <- as.factor(data$PhaseOfFlight)
data$Time <- as.factor(data$Time)
data$Place <- as.factor(data$Place)
data$HasFire <- as.factor(data$HasFire)
data$Environment <- as.factor(data$Environment)
data$Energy_absorption <- as.factor(data$Energy_absorption)





str(data)
summary(data)
#> data <- read.csv("Aerei_Final.csv")
## 'data.frame': 47 obs. of 26 variables:
## $ Airline : chr "Singapore Airlines" "British Airtours" "British Midland" "China Airlines"
## $ NumVolo : int 6 28 92 120 123 129 140 148 191 204 ...
## $ X1.terzo.lievi : int 17 36 0 0 0 4 0 0 0 3 ...
## $ X1.terzo.gravi : int 2 0 11 0 0 0 0 0 0 0 ...
## $ X1.terzo.morti : int 15 0 22 0 136 14 18 16 55 33 ...
## $ X2.terzo.lievi : int 1 30 4 5 0 5 7 1 0 25 ...
## $ X2.terzo.gravi : int 15 0 30 0 0 0 0 0 8 0 ...
## $ X2.terzo.morti : int 64 16 13 8 214 60 139 34 51 1 ...
## $ X3.terzo.lievi : int 26 10 0 5 36 24 0 8 10 29 ...
## $ X3.terzo.gravi : int 17 0 27 0 0 0 0 0 7 0 ...
## $ X3.terzo.morti : int 0 36 11 21 109 43 91 30 16 0 ...
## $ X1.meta.lievi : int 17 56 32 3 0 7 7 1 0 7 ...
## $ X1.meta.gravi : int 3 0 0 0 0 0 0 0 1 0 ...
## $ X1.meta.morti : int 41 8 34 6 226 23 95 36 79 35 ...
## $ X2.meta.lievi : int 26 20 39 7 4 26 0 7 10 46 ...
## $ X2.meta.gravi : int 31 0 0 0 0 0 0 0 14 0 ...
## $ X2.meta.morti : int 34 44 13 23 225 90 145 46 48 0 ...
## $ fonte : chr "W" "W" "W" "W" ...
## $ PhaseOfFlight : chr "Takeoff" "Takeoff" "Landing" "Landing" ...
## $ Time : chr "Night" "Day" "Night" "Day" ...
2
## $ Place : chr "Airport " "Outside" "Outside" "Airport " ...
## $ HasFire : chr "Fire" "Fire" "Fire" "Fire" ...
## $ Crushed_fuselage : int 1 1 1 1 1 1 1 1 1 1 ...
## $ Restraint_intact : int 0 0 0 1 0 0 0 0 0 0 ...
## $ Environment : chr "dangerous" "dangerous" "dangerous" "clear" ...
## $ Energy_absorption: chr "nogear" "nogear" "gear" "nogear"
                                                                            
                    
```
Some random text....

```{r}
# add a colum of # of seat for each airplain section

data$X1.third.total <- data$X1.terzo.lievi + data$X1.terzo.gravi + data$X1.terzo.morti
data$X2.third.total <- data$X2.terzo.lievi + data$X2.terzo.gravi + data$X2.terzo.morti
data$X3.third.total <- data$X3.terzo.lievi + data$X3.terzo.gravi + data$X3.terzo.morti
data$X1.half.total <- data$X1.meta.lievi + data$X1.meta.gravi + data$X1.meta.morti
data$X2.half.total <- data$X2.meta.lievi + data$X2.meta.gravi + data$X2.meta.morti

str(data)
summary(data)

# now make a colum of mortality rate for each section

data$X1.third.mortality.rate <- data$X1.terzo.morti / data$X1.third.total
data$X2.third.mortality.rate <- data$X2.terzo.morti / data$X2.third.total
data$X3.third.mortality.rate <- data$X3.terzo.morti / data$X3.third.total
data$X1.half.mortality.rate <- data$X1.meta.morti / data$X1.half.total
data$X2.half.mortality.rate <- data$X2.meta.morti / data$X2.half.total



head(data)

```
```{r}
## Plot mortality rates as scatter plots for each section
## they must be separated on the x axis by group

boxplot(data$X1.third.mortality.rate, 
  data$X2.third.mortality.rate, 
  data$X3.third.mortality.rate, 
  data$X1.half.mortality.rate, 
  data$X2.half.mortality.rate, 
  names = c("Front Third", "Middle Third", "Rear Third", "Front Half", "Rear Half"),
  main = "Mortality Rates by Section",
  ylab = "Mortality Rate",
  xlab = "Sections",
  col = c("red", "red", "red", "green", "green"))

```


```{r}
# Perform ANOVA to test if there are significant differences in mortality rates between sections (only  thirds)
mortality_data <- data.frame(
  Section = rep(c("Front Third", "Middle Third", "Rear Third"), each = nrow(data)),
  MortalityRate = c(data$X1.third.mortality.rate, data$X2.third.mortality.rate, data$X3.third.mortality.rate)
)
str(mortality_data)
mortality_data$Section <- as.factor(mortality_data$Section)

anova_result <- aov(MortalityRate ~ Section, data = mortality_data)
summary(anova_result)

library(multcomp)

test_result <- glht(anova_result, linfct = mcp(Section = "Tukey"))
summary(test_result)

#half sections
mortality_data_half <- data.frame(
  Section = rep(c("Front Half", "Rear Half"), each = nrow(data)),
  MortalityRate = c(data$X1.half.mortality.rate, data$X2.half.mortality.rate)
)
anova_result_half <- aov(MortalityRate ~ Section, data = mortality_data_half)
summary(anova_result_half)


#try using non-parametric test if ANOVA assumptions are not met
kruskal_result <- kruskal.test(MortalityRate ~ Section, data = mortality_data)
kruskal_result_half <- kruskal.test(MortalityRate ~ Section, data = mortality_data_half)
kruskal_result
kruskal_result_half

#oneway test

oneway_result <- oneway.test(MortalityRate ~ Section, data = mortality_data, var.equal = FALSE)
oneway_result_half <- oneway.test(MortalityRate ~ Section, data = mortality_data_half, var.equal = FALSE)
oneway_result
oneway_result_half

```

```{r}
# now we do the same thing but considering the "gravi" as casualties too

data$X1.casualties_rate_new <- (data$X1.terzo.morti + data$X1.terzo.gravi) / data$X1.third.total
data$X2.casualties_rate_new <- (data$X2.terzo.morti + data$X2.terzo.gravi) / data$X2.third.total
data$X3.casualties_rate_new <- (data$X3.terzo.morti + data$X3.terzo.gravi) / data$X3.third.total
data$X1.half.casualties_rate_new <- (data$X1.meta.morti + data$X1.meta.gravi) / data$X1.half.total
data$X2.half.casualties_rate_new <- (data$X2.meta.morti + data$X2.meta.gravi) / data$X2.half.total

head(data)


# Plot new casualties rates as scatter plots for each section
boxplot(data$X1.casualties_rate_new, 
  data$X2.casualties_rate_new, 
  data$X3.casualties_rate_new, 
  data$X1.half.casualties_rate_new, 
  data$X2.half.casualties_rate_new, 
  names = c("Front Third", "Middle Third", "Rear Third", "Front Half", "Rear Half"),
  main = "Casualties Rates by Section (Including Serious Injuries)",
  ylab = "Casualties Rate",
  xlab = "Sections",
  col = c("blue", "blue", "blue", "orange", "orange"))  

```


```{r}
# Perform ANOVA to test if there are significant differences in casualties rates between sections (only thirds)
casualties_data <- data.frame(
  Section = rep(c("Front Third", "Middle Third", "Rear Third"), each = nrow(data)),
  CasualtiesRate = c(data$X1.casualties_rate_new, data$X2.casualties_rate_new, data$X3.casualties_rate_new) 
)
anova_casualties_result <- aov(CasualtiesRate ~ Section, data = casualties_data)
summary(anova_casualties_result)

#half sections
casualties_data_half <- data.frame(
  Section = rep(c("Front Half", "Rear Half"), each = nrow(data)),
  CasualtiesRate = c(data$X1.half.casualties_rate_new, data$X2.half.casualties_rate_new) 
)
anova_casualties_result_half <- aov(CasualtiesRate ~ Section, data = casualties_data_half)
summary(anova_casualties_result_half) 

#try using non-parametric test if ANOVA assumptions are not met
kruskal_casualties_result <- kruskal.test(CasualtiesRate ~ Section, data = casualties_data)
kruskal_casualties_result_half <- kruskal.test(CasualtiesRate ~ Section, data = casualties_data_half)
kruskal_casualties_result
kruskal_casualties_result_half

#oneway test
oneway_casualties_result <- oneway.test(CasualtiesRate ~ Section, data = casualties_data, var.equal = FALSE)
oneway_casualties_result_half <- oneway.test(CasualtiesRate ~ Section, data = casualties_data_half, var.equal = FALSE)
oneway_casualties_result
oneway_casualties_result_half


```



```{r}
# t.test only the fisrts third against the rears third
t_test_result <- t.test(data$X1.third.mortality.rate, data$X3.third.mortality.rate, var.equal = FALSE)
t_test_result

#use non-parametric test if t-test assumptions are not met
wilcox_test_result <- wilcox.test(data$X1.third.mortality.rate, data$X3.third.mortality.rate)
wilcox_test_result

```



```{r}
# now we want to test if there is a difference in mortality rate based on the other variables
# PhaseOfFlight, Time, Place, HasFire
# using the total mortality rate, only deaths, do this for evwey third

out <- lm(data$X1.third.mortality.rate ~ data$PhaseOfFlight + data$Time + data$Place + data$HasFire + data$Environment + data$Energy_absorption + data$Crushed_fuselage + data$Restraint_intact)
summary(out)

##simpify the model susins stepwise regression

out_simple <- step(out)
summary(out_simple)

out_2 <- lm(data$X2.third.mortality.rate ~ data$PhaseOfFlight + data$Time + data$Place + data$HasFire + data$Environment + data$Energy_absorption + data$Crushed_fuselage + data$Restraint_intact)
summary(out_2)

out_2_simple <- step(out_2)
summary(out_2_simple)

out_3 <- lm(data$X3.third.mortality.rate ~ data$PhaseOfFlight + data$Time + data$Place + data$HasFire + data$Environment + data$Energy_absorption + data$Crushed_fuselage + data$Restraint_intact)
summary(out_3)

out_3_simple <- step(out_3)
summary(out_3_simple)


#plot the mortality rates based on HasFire
boxplot(data$X1.third.mortality.rate ~ data$HasFire,
        main = "Front Third Mortality Rate by Fire Presence",
        xlab = "Has Fire",
        ylab = "Mortality Rate",
        col = c("yellow", "red"))

boxplot(data$X2.third.mortality.rate ~ data$HasFire,
        main = "Middle Third Mortality Rate by Fire Presence",
        xlab = "Has Fire",
        ylab = "Mortality Rate",
        col = c("yellow", "red")) 

boxplot(data$X3.third.mortality.rate ~ data$HasFire,
        main = "Rear Third Mortality Rate by Fire Presence",
        xlab = "Has Fire",
        ylab = "Mortality Rate",
        col = c("yellow", "red")) 

#and full plane mortality rate
data$Total.mortality.rate <- (data$X1.terzo.morti + data$X2.terzo.morti + data$X3.terzo.morti) / (data$X1.third.total + data$X2.third.total + data$X3.third.total)

boxplot(data$Total.mortality.rate ~ data$HasFire,
        main = "Total Mortality Rate by Fire Presence",
        xlab = "Has Fire",
        ylab = "Mortality Rate",
        col = c("yellow", "red"))

# now for the halves
out_half_1 <- lm(data$X1.half.mortality.rate ~ data$PhaseOfFlight + data$Time + data$Place + data$HasFire + data$Environment + data$Energy_absorption + data$Crushed_fuselage + data$Restraint_intact)
summary(out_half_1)

#simplify the model using stepwise regression
out_half_1_simple <- step(out_half_1 , direction = "both", trace = 0)
summary(out_half_1_simple)


out_half_2 <- lm(data$X2.half.mortality.rate ~ data$PhaseOfFlight + data$Time + data$Place + data$HasFire + data$Environment + data$Energy_absorption + data$Crushed_fuselage + data$Restraint_intact)
summary(out_half_2)

#simplify the model using stepwise regression
out_half_2_simple <- step(out_half_2 , direction = "both", trace = 0)
summary(out_half_2_simple)


#boxplots for halves
boxplot(data$X1.half.mortality.rate ~ data$HasFire,
        main = "Front Half Mortality Rate by Fire Presence",
        xlab = "Has Fire",    
        ylab = "Mortality Rate",
        col = c("lightgreen", "darkgreen"))

boxplot(data$X2.half.mortality.rate ~ data$HasFire,
        main = "Rear Half Mortality Rate by Fire Presence",
        xlab = "Has Fire",
        ylab = "Mortality Rate",
        col = c("lightgreen", "darkgreen")) 



############################################################

#same analysis but using glm with binomial family

out_glm <- glm(cbind(data$X1.terzo.morti, data$X1.third.total - data$X1.terzo.morti) ~ data$PhaseOfFlight + data$Time + data$Place + data$HasFire + data$Environment + data$Energy_absorption + data$Crushed_fuselage + data$Restraint_intact, family = binomial)

summary(out_glm)
out_glm_simple <- step(out_glm, direction = "both", trace = 0)
summary(out_glm_simple)

out_glm_2 <- glm(cbind(data$X2.terzo.morti, data$X2.third.total - data$X2.terzo.morti) ~ data$PhaseOfFlight + data$Time + data$Place + data$HasFire + data$Environment + data$Energy_absorption + data$Crushed_fuselage + data$Restraint_intact, family = binomial)

summary(out_glm_2)
out_glm_2_simple <- step(out_glm_2, direction = "both", trace = 0)
summary(out_glm_2_simple)

out_glm_3 <- glm(cbind(data$X3.terzo.morti, data$X3.third.total - data$X3.terzo.morti) ~ data$PhaseOfFlight + data$Time + data$Place + data$HasFire + data$Environment + data$Energy_absorption + data$Crushed_fuselage + data$Restraint_intact, family = binomial)

summary(out_glm_3)
out_glm_3_simple <- step(out_glm_3, direction = "both", trace = 0)
summary(out_glm_3_simple)

# now for the halves
out_glm_half_1 <- glm(cbind(data$X1.meta.morti, data$X1.half.total - data$X1.meta.morti) ~ data$PhaseOfFlight + data$Time + data$Place + data$HasFire + data$Environment + data$Energy_absorption + data$Crushed_fuselage + data$Restraint_intact, family = binomial)
summary(out_glm_half_1)

out_glm_half_1_simple <- step(out_glm_half_1 , direction = "both", trace = 0)
summary(out_glm_half_1_simple)

out_glm_half_2 <- glm(cbind(data$X2.meta.morti, data$X2.half.total - data$X2.meta.morti) ~ data$PhaseOfFlight + data$Time + data$Place + data$HasFire + data$Environment + data$Energy_absorption + data$Crushed_fuselage + data$Restraint_intact, family = binomial)
summary(out_glm_half_2)
out_glm_half_2_simple <- step(out_glm_half_2 , direction = "both", trace = 0)
summary(out_glm_half_2_simple)



```


```{r}
#same lm analysis but considering casualties (deaths + serious injuries)
out_casualties <- lm(data$X1.casualties_rate_new ~ data$PhaseOfFlight + data$Time + data$Place + data$HasFire + data$Environment + data$Energy_absorption + data$Crushed_fuselage + data$Restraint_intact)
summary(out_casualties)

#simpify the model susins stepwise regression 
out_casualties_simple <- step(out_casualties, direction = "both", trace = 0)
summary(out_casualties_simple)

out_casualties_2 <- lm(data$X2.casualties_rate_new ~ data$PhaseOfFlight + data$Time + data$Place + data$HasFire + data$Environment + data$Energy_absorption + data$Crushed_fuselage + data$Restraint_intact)
summary(out_casualties_2)

out_casualties_2_simple <- step(out_casualties_2, direction = "both", trace = 0)
summary(out_casualties_2_simple)

out_casualties_3 <- lm(data$X3.casualties_rate_new ~ data$PhaseOfFlight + data$Time + data$Place + data$HasFire + data$Environment + data$Energy_absorption + data$Crushed_fuselage + data$Restraint_intact)
summary(out_casualties_3)

out_casualties_3_simple <- step(out_casualties_3, direction = "both", trace = 0)
summary(out_casualties_3_simple)

# now for the halves
out_casualties_half_1 <- lm(data$X1.half.casualties_rate_new ~ data$PhaseOfFlight + data$Time + data$Place + data$HasFire + data$Environment + data$Energy_absorption + data$Crushed_fuselage + data$Restraint_intact)
summary(out_casualties_half_1)

#simplify the model using stepwise regression
out_casualties_half_1_simple <- step(out_casualties_half_1 , direction = "both", trace = 0)
summary(out_casualties_half_1_simple)


out_casualties_half_2 <- lm(data$X2.half.casualties_rate_new ~ data$PhaseOfFlight + data$Time + data$Place + data$HasFire + data$Environment + data$Energy_absorption + data$Crushed_fuselage + data$Restraint_intact)
summary(out_casualties_half_2)
#simplify the model using stepwise regression
out_casualties_half_2_simple <- step(out_casualties_half_2 , direction = "both", trace = 0)
summary(out_casualties_half_2_simple) 




```



```{r}
# now we wnar to study the correlation between the different sections mortality rates
correlation_matrix <- data.frame(
  Front_Third = data$X1.third.mortality.rate,
  Middle_Third = data$X2.third.mortality.rate,
  Rear_Third = data$X3.third.mortality.rate,
  Front_Half = data$X1.half.mortality.rate,
  Rear_Half = data$X2.half.mortality.rate
  
)
correlation_results <- cor(correlation_matrix, use = "complete.obs")
correlation_results

## cor plot
library(corrplot)
corrplot(correlation_results, method = "ellipse", type = "upper", tl.col = "black", tl.srt = 45)

## now study the correlation between the other variables and the total mortality rate
correlation_matrix_2 <- data.frame(
  Total_Mortality_Rate = data$Total.mortality.rate,
  PhaseOfFlight = as.numeric(as.factor(data$PhaseOfFlight)),
  Time = as.numeric(as.factor(data$Time)),
  Place = as.numeric(as.factor(data$Place)),
  HasFire = as.numeric(as.factor(data$HasFire)),
  Environment = as.numeric(as.factor(data$Environment)),
  Energy_absorption = as.numeric(as.factor(data$Energy_absorption)),
  Crushed_fuselage = as.numeric(as.factor(data$Crushed_fuselage)),
  Restraint_intact = as.numeric(as.factor(data$Restraint_intact))
)
correlation_results_2 <- cor(correlation_matrix_2, use = "complete.obs")
correlation_results_2

corrplot(correlation_results_2, method = "ellipse", type = "upper", tl.col = "black", tl.srt = 45)



```

# Data Description & Analysis

Firstly, let's take a look at the data gathered,in particular its structure and meaning:

```{r load data}
accident_data <- read.csv("AllCREEP_cleaned_eng.csv")
str(accident_data)
#
#> str(accident_data)
#'data.frame':   47 obs. of  26 variables:
# $ Airline          : chr  "singapore airlines" "british airtours" "british midland" "china airlines" ...
# $ FlightNum        : int  6 28 92 120 123 129 140 148 191 204 ...
# $ X1.third.minor   : int  17 36 0 0 0 4 0 0 0 3 ...
# $ X1.third.major   : int  2 0 11 0 0 0 0 0 0 0 ...
# $ X1.third.dead    : int  15 0 22 0 136 14 18 16 55 33 ...
# $ X2.third.minor   : int  1 30 4 5 0 5 7 1 0 25 ...
# $ X2.third.major   : int  15 0 30 0 0 0 0 0 8 0 ...
# $ X2.third.dead    : int  64 16 13 8 214 60 139 34 51 1 ...
# $ X3.third.minor   : int  26 10 0 5 36 24 0 8 10 29 ...
# $ X3.third.major   : int  17 0 27 0 0 0 0 0 7 0 ...
# $ X3.third.dead    : int  0 36 11 21 109 43 91 30 16 0 ...
# $ X1.half.minor    : int  17 56 32 3 0 7 7 1 0 7 ...
# $ X1.half.major    : int  3 0 0 0 0 0 0 0 1 0 ...
# $ X1.half.dead     : int  41 8 34 6 226 23 95 36 79 35 ...
# $ X2.half.minor    : int  26 20 39 7 4 26 0 7 10 46 ...
# $ X2.half.major    : int  31 0 0 0 0 0 0 0 14 0 ...
# $ X2.half.dead     : int  34 44 13 23 225 90 145 46 48 0 ...
# $ DataOrigin       : chr  "w" "w" "w" "w" ...
# $ PhaseOfFlight    : chr  "takeoff" "takeoff" "landing" "landing" ...
# $ Time             : chr  "night" "day" "night" "day" ...
# $ Place            : chr  "airport" "outside" "outside" "airport" ...
# $ HasFire          : chr  "fire" "fire" "fire" "fire" ...
# $ CrushedFuselage : int  1 1 1 1 1 1 1 1 1 1 ...
# $ RestraintIntact : int  0 0 0 1 0 0 0 0 0 0 ...
# $ Environment      : chr  "dangerous" "dangerous" "dangerous" "clear" ...
# $ EnergyAbsorption: chr  "nogear" "nogear" "gear" "nogear" ...
#> 
#
```

The data consists of 47 observations (aircraft accidents) and 26 variables, that are the focus of our analysis. We want to remark that the data consists only of aircraft accidents where there were at least one survival.

The variables consists of:

- **Airline & FlightNum** : identification of the crash, usefull to retrive more information about the accident if needed

- **X variables** : these variables represent the number and the type of casualties in each section of the airplane. The plane is divided in two different ways : in thirds and in halves. For each section, there are three variables representing the number of minor injuries, major injuries and deaths for that section. The format is the following : `X{section}.{part}.{type of injury}`, where section is 1,2,3 for the thirds and 1,2 for the halves, part is "third" or "half", and type of injury is "minor", "major" or "dead".

- **DataOrigin** : source of the data, [ to complete ]

- **PhaseOfFlight** : phase of flight when the accident happened, The possible values are "takeoff" or "landing", landing includes possible emergency landings.

- **Time** : time of the day when the accident happened, possible values are "day" or "night". This is intended to signal if natural light was present or not.
- **Place** : location where the accident happened, possible values are "airport" or "outside". "Airport" means that the accident happened on the runway or in the immediate vicinity of the airport, "outside" means that the accident happened far from any airport.

- **HasFire**  : indicates if there was a fire during or after the accident, possible values are "fire" or "no-fire".

Now we include also 4 other variables useful to determine the severity of an aircraft accident:

- **CrushedFuselage** : indicates if the fuselage was crushed during the accident, possible values are 1 (yes) or 0 (no).
- **RestraintIntact** : indicates if the restraint system was intact, and thus effective at retain the passengers, during the accident, possible values are 1 (yes) or 0 (no).

- **Environment** : indicates the environmental conditions during the accident, possible values are "clear" or "dangerous".

- **EnergyAbsorption** : indicates if the landing gear was developed to absorb energy during the accident, if not the airplane is considerd to have done a "belly landing".


To start the analysis, let's prepare the data by adding some useful variables, and trasform some variables into a more manageable format.
```{r data preparation}
# add a colum of # of passengers for each airplain section
accident_data <- within(accident_data, {
  X2.half.total  <- X2.half.minor  + X2.half.major  + X2.half.dead
  X1.half.total  <- X1.half.minor  + X1.half.major  + X1.half.dead
  X3.third.total <- X3.third.minor + X3.third.major + X3.third.dead
  X2.third.total <- X2.third.minor + X2.third.major + X2.third.dead
  X1.third.total <- X1.third.minor + X1.third.major + X1.third.dead
})

# now we transform the variables that are categorical into factors
factor_cols <- c("PhaseOfFlight", "Time", "Place", "HasFire", "Environment", 
                 "EnergyAbsorption", "CrushedFuselage", "RestraintIntact")

accident_data[factor_cols] <- lapply(accident_data[factor_cols], as.factor)
```

Now we can add a new variable that will be the center of our analysis : the mortality rate for each section of the airplane. The mortality rate is defined as the number of deaths divided by the total number of passengers in that section. Note that some of this aircraft where not full, so there might be sections with very few passengers or possibly empty. Thats why we are considering rates per total passengers instead of per number of seats.
```{r add mortality rates}
accident_data <- within(accident_data, {
  X1.third.mortality.rate <- X1.third.dead / X1.third.total
  X2.third.mortality.rate <- X2.third.dead / X2.third.total
  X3.third.mortality.rate <- X3.third.dead / X3.third.total
  X1.half.mortality.rate  <- X1.half.dead  / X1.half.total
  X2.half.mortality.rate  <- X2.half.dead  / X2.half.total
})
```

## Exploratory Data Analysis

The first step we can do is to plot some feature of the data to have a visual understanding of it. We might get some useful insights and also check for possible outliers or anomalies in the data.

```{r simple boxplots}
with(accident_data, {
  boxplot(X1.third.mortality.rate, 
          X2.third.mortality.rate, 
          X3.third.mortality.rate, 
          X1.half.mortality.rate, 
          X2.half.mortality.rate, 
          names = c("Front Third", "Middle Third", "Rear Third", "Front Half", "Rear Half"),
          main = "Mortality Rates by Section",
          ylab = "Mortality Rate",
          xlab = "Sections",
          col = c("red", "red", "red", "green", "green"),
          cex.axis = 0.9)
})
```

Intestingly, there seems to be a general trend of lower mortality rates the more rear the section is. This is seen in both the third and half divisions of the airplane.

Now we can try to test the kind of population distribution of the mortality rates, to see if they follow a normal distribution or not. This will help us choose the right statistical tests for our analysis.

We will try the standard Shapiro-Wilk test for normality.

```{r normality tests}
# shapiro test for normality, p-value extrated and lableled
shapiro_results <- data.frame(
  Section = c("Front Third", "Middle Third", "Rear Third", "Front Half", "Rear Half"),
  P_Value = c(
    shapiro.test(accident_data$X1.third.mortality.rate)$p.value,
    shapiro.test(accident_data$X2.third.mortality.rate)$p.value,
    shapiro.test(accident_data$X3.third.mortality.rate)$p.value,
    shapiro.test(accident_data$X1.half.mortality.rate)$p.value,
    shapiro.test(accident_data$X2.half.mortality.rate)$p.value
  )
)
shapiro_results
```

All of the p-values are far below the standard threshold of 0.05, indicating that we can reject the null hypothesis of normality for all sections. This means that the mortality rates do not follow a normal distribution, let's try to plot a histogram for one of the sections to visualize an aproximately of the distribution.

```{r istrogram distribution, fig.height=4, fig.width=8}
par(mfrow = c(1, 2))

colums <- c("X1.third.mortality.rate", "X3.third.mortality.rate") 

for( c in colums){
  hist(accident_data[[c]],
       main = paste("Histogram of", gsub("\\.", " ", c)),
       xlab = "Mortality Rate",
       col = "lightblue",
       border = "black",
       breaks = 10)
}
```

There seems that there is no clear distribution pattern, but we can se that the data is more concentrated  at either ends of the mortality rate values, expecially for the front third section.

## Statistical Analysis

Now that we have a better understanding of the data, we can proceed with the statistical analysis to test if there are significant differences in mortality rates between the different sections of the airplane. Also we will try to see if there is some correlation between the mortality rates and the other variables in the dataset.

### Analysis of differences in mortality rates between sections

Since the distribution of the mortality rates is not normal, we will not use test that assume normality, like t-tests or ANOVA.
For this reason, we will use the Kruskal-Wallis test, which is a non-parametric test to compare the distribution of two or more groups. The null hypothesis of the Kruskal-Wallis test is that all the populations have the same distribution.

```{r analysis of population differences}
# Kruskal-Wallis test for differences in mortality rates between sections (only thirds)
mortality_data_third <- data.frame(
  SectionThird = rep(c("Front Third", "Middle Third", "Rear Third"), each = nrow(accident_data)),
  MortalityRateThird = c(accident_data$X1.third.mortality.rate, 
                    accident_data$X2.third.mortality.rate, 
                    accident_data$X3.third.mortality.rate)
)

kruskal_result_third <- kruskal.test(MortalityRateThird ~ SectionThird, data = mortality_data_third)
kruskal_result_third

# Kruskal-Wallis test for differences in mortality rates between sections (only halves)
mortality_data_half <- data.frame(
  SectionHalf = rep(c("Front Half", "Rear Half"), each = nrow(accident_data)),
  MortalityRateHalf = c(accident_data$X1.half.mortality.rate, 
                    accident_data$X2.half.mortality.rate)
)

kruskal_result_half <- kruskal.test(MortalityRateHalf ~ SectionHalf, data = mortality_data_half)
kruskal_result_half
```

[Interpretation of results to be added here, OR later in the report]

### Modeling the mortality rates
Now we will try to fit a model to possibly discover the effect of the other variables on the mortality rates of the different sections. The focus is not specifically on prediction, but rather on trying to find other significant difference between the sections. Now the questions are:

- Are there variable that significantly affect the mortality rate?

- If so, do they affect differently the various sections of the airplane?

We will try to fit a Generalized Linear Model (GLM) with binomial family, since the mortality rate is a proportion (number of deaths / total number of passengers). Unfortunately, for the nature of the data, the assumptions of the binomial distribution are not met, since the passenger's deaths are not independent events. To account for this, we will use the quasibinomial family, which is a more flexible version of the binomial family that allows for overdispersion and corrects the standard errors and p-values accordingly.  

```{r test glm}
### test glm with binomial family to see the effect of other variables on mortality rate for each section

## also print the varius confidence intervals for the coefficients, and trasform them into odds ratios
glm_third_1 <- glm(cbind(accident_data$X1.third.dead, accident_data$X1.third.total - accident_data$X1.third.dead) ~ 
                     #accident_data$PhaseOfFlight + 
                     #accident_data$Time + 
                     #accident_data$Place + 
                     accident_data$HasFire + 
                     #accident_data$Environment + 
                     accident_data$EnergyAbsorption + 
                     accident_data$CrushedFuselage + 
                     accident_data$RestraintIntact, 
                   family = quasibinomial(link = "logit"))

glm_third_2 <- glm(cbind(accident_data$X2.third.dead, accident_data$X2.third.total - accident_data$X2.third.dead) ~ 
                     #accident_data$PhaseOfFlight + 
                     #accident_data$Time + 
                     #accident_data$Place + 
                     accident_data$HasFire + 
                     #accident_data$Environment + 
                     accident_data$EnergyAbsorption + 
                     accident_data$CrushedFuselage + 
                     accident_data$RestraintIntact, 
                   family = quasibinomial(link = "logit"))

glm_third_3 <- glm(cbind(accident_data$X3.third.dead, accident_data$X3.third.total - accident_data$X3.third.dead) ~ 
                     #accident_data$PhaseOfFlight + 
                     #accident_data$Time + 
                     #accident_data$Place + 
                     accident_data$HasFire + 
                     #accident_data$Environment + 
                     accident_data$EnergyAbsorption + 
                     accident_data$CrushedFuselage + 
                     accident_data$RestraintIntact, 
                   family = quasibinomial(link = "logit"))

# now try to model the total mortality rate

glm_total <- glm(cbind(accident_data$X1.third.dead + accident_data$X2.third.dead + accident_data$X3.third.dead, 
                       accident_data$X1.third.total + accident_data$X2.third.total + accident_data$X3.third.total - 
                         (accident_data$X1.third.dead + accident_data$X2.third.dead + accident_data$X3.third.dead)) ~ 
                   #accident_data$PhaseOfFlight + 
                   #accident_data$Time + 
                   #accident_data$Place + 
                   accident_data$HasFire + 
                   #accident_data$Environment + 
                   accident_data$EnergyAbsorption + 
                   accident_data$CrushedFuselage + 
                   accident_data$RestraintIntact, 
                 family = quasibinomial(link = "logit"))

```



```{r glm summaries}
summary(glm_third_1)
summary(glm_third_2)
summary(glm_third_3)
summary(glm_total)
```


```{r}
x <- rnorm(1000)
hist(x)
```

# Analysis

# Results 

# Conclusions
