---
title: "Plane Crash Analysis: There exist such a thing as a safest seat?"
subtitle: "Inferential Statistics 2025/2026"
author: "Barbato Alberto, Naldoni Martina"
date: "`r Sys.Date()`"
output: 
  pdf_document: 
    toc: true
    fig_height: 6
    number_sections: true
editor_options: 
  markdown: 
    wrap: 80
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Introduction

The Aviation industry is one of the safest in the world. The systems that it has in place to learn lessons from accidents and make sure they never happen again is well-established and highly regarded. This has made travelling by plane the safest way to travel. 
https://flyfright.com/plane-crash-statistics/#:~:text=Based%20on%20statistics%20from%202015,unharmed%2C%20injured%2C%20or%20killed

But perhaps due to these high safety standards, when an aircraft accident happens it makes headlines all over the world. Anxious passengers fear it's going to happen to them as well and they can't help but ask themselves if there is something they could do to have a safer flight. 

One of the most frequent question that gets asked is : is there a part of the plane that is "safer" than other parts? Can the seating location make a difference in an accident? 
There are many articles where experts in the field give their opinions, and most answer that "yes, there are areas that give a higher chance of survival in case of an aircraft accident". 
But there are also studies that suggest that there is no safest seats on an airplane. 

So, who is right? We are going to try to answer this question using statistics.

```{r load data}

data <- read.csv("Aerei_Final.csv")

str(data)
summary(data)
#> data <- read.csv("Aerei_Final.csv")
#> 
#> str(data)
#'data.frame':   47 obs. of  21 variables:
# $ NumVolo         : int  6 28 92 120 123 129 140 148 191 204 ...
# $ X1.terzo.lievi  : int  17 36 0 0 0 4 0 0 0 3 ...
# $ X1.terzo.gravi  : int  2 0 11 0 0 0 0 0 0 0 ...
# $ X1.terzo.morti  : int  15 0 22 0 136 14 18 16 55 33 ...
# $ X2.terzo.lievi  : int  1 30 4 5 0 5 7 1 0 25 ...
# $ X2.terzo.gravi  : int  15 0 30 0 0 0 0 0 8 0 ...
# $ X2.terzo.morti  : int  64 16 13 8 214 60 139 34 51 1 ...
# $ X3.terzo.lievi  : int  26 10 0 5 36 24 0 8 10 29 ...
# $ X3.terzo.gravi  : int  17 0 27 0 0 0 0 0 7 0 ...
# $ X2.terzo.morti.1: int  0 36 11 21 109 43 91 30 16 0 ...
# $ X1.meta.lievi   : int  17 56 32 3 0 7 7 1 0 7 ...
# $ X1.meta.gravi   : int  3 0 0 0 0 0 0 0 1 0 ...
# $ X1.meta.morti   : int  41 8 34 6 226 23 95 36 79 35 ...
# $ X2.meta.lievi   : int  26 20 39 7 4 26 0 7 10 46 ...
# $ X2.meta.gravi   : int  31 0 0 0 0 0 0 0 14 0 ...
# $ X2.meta.morti   : int  34 44 13 23 225 90 145 46 48 0 ...
# $ fonte           : chr  "W" "W" "W" "W" ...
# $ PhaseOfFlight   : chr  "Takeoff" "Takeoff" "Landing" "Landing" ...
# $ Time            : chr  "Night" "Day" "Night" "Day" ...
# $ Place           : chr  "Airport " "Outside" "Outside" "Airport " ...
# $ HasFire         : chr  "Fire" "Fire" "Fire" "Fire" ...
#> summary(data)
#    NumVolo     X1.terzo.lievi   X1.terzo.gravi   X1.terzo.morti  
# Min.   :   6   Min.   :  0.00   Min.   : 0.000   Min.   :  0.00  
# 1st Qu.: 227   1st Qu.:  0.00   1st Qu.: 0.000   1st Qu.:  1.50  
# Median : 812   Median :  4.00   Median : 0.000   Median : 11.00  
# Mean   :1591   Mean   : 12.87   Mean   : 2.553   Mean   : 16.15  
# 3rd Qu.:1603   3rd Qu.: 17.00   3rd Qu.: 4.000   3rd Qu.: 19.00  
# Max.   :9642   Max.   :141.00   Max.   :16.000   Max.   :136.00  
# X2.terzo.lievi   X2.terzo.gravi   X2.terzo.morti   X3.terzo.lievi 
# Min.   :  0.00   Min.   : 0.000   Min.   :  0.00   Min.   :  0.0  
# 1st Qu.:  1.00   1st Qu.: 0.000   1st Qu.:  2.00   1st Qu.:  3.0  
# Median :  7.00   Median : 2.000   Median : 12.00   Median : 10.0  
# Mean   : 22.11   Mean   : 4.638   Mean   : 28.62   Mean   : 20.4  
# 3rd Qu.: 26.50   3rd Qu.: 8.000   3rd Qu.: 31.00   3rd Qu.: 27.0  
# Max.   :174.00   Max.   :30.000   Max.   :214.00   Max.   :142.0  
# X3.terzo.gravi   X2.terzo.morti.1 X1.meta.lievi    X1.meta.gravi  
# Min.   : 0.000   Min.   :  0.00   Min.   :  0.00   Min.   : 0.00  
# 1st Qu.: 0.000   1st Qu.:  1.50   1st Qu.:  1.50   1st Qu.: 0.00  
# Median : 0.000   Median :  8.00   Median : 11.00   Median : 0.00  
# Mean   : 3.872   Mean   : 19.83   Mean   : 29.23   Mean   : 4.34  
# 3rd Qu.: 4.500   3rd Qu.: 24.50   3rd Qu.: 40.50   3rd Qu.: 7.00  
# Max.   :27.000   Max.   :113.00   Max.   :221.00   Max.   :34.00  
# X1.meta.morti    X2.meta.lievi    X2.meta.gravi    X2.meta.morti   
# Min.   :  0.00   Min.   :  0.00   Min.   : 0.000   Min.   :  0.00  
# 1st Qu.:  5.00   1st Qu.:  6.50   1st Qu.: 0.000   1st Qu.:  3.50  
# Median : 20.00   Median : 14.00   Median : 0.000   Median : 15.00  
# Mean   : 28.57   Mean   : 28.98   Mean   : 5.234   Mean   : 36.53  
# 3rd Qu.: 35.00   3rd Qu.: 37.50   3rd Qu.: 7.500   3rd Qu.: 45.00  
# Max.   :226.00   Max.   :184.00   Max.   :31.000   Max.   :225.00  
#    fonte           PhaseOfFlight          Time              Place          
# Length:47          Length:47          Length:47          Length:47         
# Class :character   Class :character   Class :character   Class :character  
# Mode  :character   Mode  :character   Mode  :character   Mode  :character  
                                                                            
                    
```
Some random text....

```{r}
# add a colum of # of seat for each airplain section

data$X1.third.total <- data$X1.terzo.lievi + data$X1.terzo.gravi + data$X1.terzo.morti
data$X2.third.total <- data$X2.terzo.lievi + data$X2.terzo.gravi + data$X2.terzo.morti
data$X3.third.total <- data$X3.terzo.lievi + data$X3.terzo.gravi + data$X3.terzo.morti
data$X1.half.total <- data$X1.meta.lievi + data$X1.meta.gravi + data$X1.meta.morti
data$X2.half.total <- data$X2.meta.lievi + data$X2.meta.gravi + data$X2.meta.morti

# now make a colum of mortality rate for each section

data$X1.third.mortality.rate <- data$X1.terzo.morti / data$X1.third.total
data$X2.third.mortality.rate <- data$X2.terzo.morti / data$X2.third.total
data$X3.third.mortality.rate <- data$X3.terzo.morti / data$X3.third.total
data$X1.half.mortality.rate <- data$X1.meta.morti / data$X1.half.total
data$X2.half.mortality.rate <- data$X2.meta.morti / data$X2.half.total



head(data)

```
```{r}
## Plot mortality rates as scatter plots for each section
## they must be separated on the x axis by group

boxplot(data$X1.third.mortality.rate, 
  data$X2.third.mortality.rate, 
  data$X3.third.mortality.rate, 
  data$X1.half.mortality.rate, 
  data$X2.half.mortality.rate, 
  names = c("Front Third", "Middle Third", "Rear Third", "Front Half", "Rear Half"),
  main = "Mortality Rates by Section",
  ylab = "Mortality Rate",
  xlab = "Sections",
  col = c("red", "red", "red", "green", "green"))

```


```{r}
# Perform ANOVA to test if there are significant differences in mortality rates between sections (only  thirds)
mortality_data <- data.frame(
  Section = rep(c("Front Third", "Middle Third", "Rear Third"), each = nrow(data)),
  MortalityRate = c(data$X1.third.mortality.rate, data$X2.third.mortality.rate, data$X3.third.mortality.rate)
)
anova_result <- aov(MortalityRate ~ Section, data = mortality_data)
summary(anova_result)

#half sections
mortality_data_half <- data.frame(
  Section = rep(c("Front Half", "Rear Half"), each = nrow(data)),
  MortalityRate = c(data$X1.half.mortality.rate, data$X2.half.mortality.rate)
)
anova_result_half <- aov(MortalityRate ~ Section, data = mortality_data_half)
summary(anova_result_half)


#try using non-parametric test if ANOVA assumptions are not met
kruskal_result <- kruskal.test(MortalityRate ~ Section, data = mortality_data)
kruskal_result_half <- kruskal.test(MortalityRate ~ Section, data = mortality_data_half)
kruskal_result
kruskal_result_half

```

```{r}
# now we do the same thing but considering the "gravi" as casualties too

data$X1.casualties_rate_new <- (data$X1.terzo.morti + data$X1.terzo.gravi) / data$X1.third.total
data$X2.casualties_rate_new <- (data$X2.terzo.morti + data$X2.terzo.gravi) / data$X2.third.total
data$X3.casualties_rate_new <- (data$X3.terzo.morti + data$X3.terzo.gravi) / data$X3.third.total
data$X1.half.casualties_rate_new <- (data$X1.meta.morti + data$X1.meta.gravi) / data$X1.half.total
data$X2.half.casualties_rate_new <- (data$X2.meta.morti + data$X2.meta.gravi) / data$X2.half.total

head(data)


# Plot new casualties rates as scatter plots for each section
boxplot(data$X1.casualties_rate_new, 
  data$X2.casualties_rate_new, 
  data$X3.casualties_rate_new, 
  data$X1.half.casualties_rate_new, 
  data$X2.half.casualties_rate_new, 
  names = c("Front Third", "Middle Third", "Rear Third", "Front Half", "Rear Half"),
  main = "Casualties Rates by Section (Including Serious Injuries)",
  ylab = "Casualties Rate",
  xlab = "Sections",
  col = c("blue", "blue", "blue", "orange", "orange"))  

```


```{r}
# Perform ANOVA to test if there are significant differences in casualties rates between sections (only thirds)
casualties_data <- data.frame(
  Section = rep(c("Front Third", "Middle Third", "Rear Third"), each = nrow(data)),
  CasualtiesRate = c(data$X1.casualties_rate_new, data$X2.casualties_rate_new, data$X3.casualties_rate_new) 
)
anova_casualties_result <- aov(CasualtiesRate ~ Section, data = casualties_data)
summary(anova_casualties_result)

#half sections
casualties_data_half <- data.frame(
  Section = rep(c("Front Half", "Rear Half"), each = nrow(data)),
  CasualtiesRate = c(data$X1.half.casualties_rate_new, data$X2.half.casualties_rate_new) 
)
anova_casualties_result_half <- aov(CasualtiesRate ~ Section, data = casualties_data_half)
summary(anova_casualties_result_half) 

#try using non-parametric test if ANOVA assumptions are not met
kruskal_casualties_result <- kruskal.test(CasualtiesRate ~ Section, data = casualties_data)
kruskal_casualties_result_half <- kruskal.test(CasualtiesRate ~ Section, data = casualties_data_half)
kruskal_casualties_result
kruskal_casualties_result_half



```


# Data Description

This is an example of r-chuch. You can instrt the R code, Knit-it and your report
is compiled in pdf.

```{r}
x <- rnorm(1000)
hist(x)
```

# Analysis

# Results 

# Conclusions
